Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-02-24 writing/manuscript.Rmd
finished first complete draft of the manuscript

---
title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection. We then leverage the estimated effects to predict when, how, and where mammals will move in BC throughout the 21^st^ century. We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts. ***MISSING RESULTS AND CONCLUSIONS***"
```

@@

\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, habitat selection

@@

* A review of climate change effects on marine mammals in United States waters: Past predictions, observed impacts, current research and conservation imperatives https://doi.org/10.1016/j.ecochg.2022.100054
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3

@@

<!-- movement is a convenient way to respond to change -->

\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008], including escaping predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024], searching for resources [@charnov_optimal_1976; @mezzini_how_2024], and habitat selection [@kacelnik_ideal_1992; @winter_forecasting_2024]. However, movement efficiency depends on ambient temperature, since it impacts the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Additionally, temperature also affects movement indirectly by affecting other drivers of movement behavior, including physiology [@abram_behavioural_2017], fitness [@schmidt_interplay_2020; @berger_climate_2018], trophic interactions [@hegel_interacting_2010], and the availability and value of resources [e.g., water and heat, see: @elmore_implications_2017; @attias_effects_2018].

<!-- changes in temperature cause changes in movement -->

Changes in temperature affect not only whether and how fast animals move but also how they move [i.e., speed and tortuosity; their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023]. Thus, temperature affects not only an animal's energetic expenditure but also many other consequences of movement, including encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and the occurrence of human-wildlife conflict [@weststrate_how_2024]. These changes are exacerbated at extreme temperatures, as animals often alter their movement behaviors substantially when temperatures are excessively cold or hot [@wilson_stenotherms_2001; @berger_climate_2018; @arechavala-lopez_common_2019; @giroux_activity_2023]. But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes above or below their ideal range [@giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024]. However, this grater adaptability is tested when generalists are subject to changes that are more severe [@mccain_assessing_2019], affecting many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or too rapid to adapt to within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021]. Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations do not react to changes in environmental conditions [@sawyer_migratory_2019] or are physiologically incapable [@sherwood_adaptability_2010; @williams_range_2018].

@@

This uncertainty in both future conditions and their consequences on ecosystems and populations presents a need for a comprehensive understanding of the effects of temperature on the movement of animal populations as well as the consequences this will have on landscapes, biodiversity, and their adaptability and resiliency. Recent work has documented climate-change-related changes mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023]; thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024]. In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes will worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with growing anthopogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming, phenological shifts, and more frequent and intense extreme events, including forest fires [@zurowski_summer_2023]. In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], we stress that decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future. Failing to understand how changes in temperature will affect mammalian movement behavior and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides an analysis of the effects of spatiotemporally proximal temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over two decades of telemetry data over a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in the current century under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017]. We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts while disentangling the direct and indirect effects of temperature on the movement behavior of terrestrial mammals.

```{r bc-map, fig.cap="GPS telemetry data for the six species in this study."}
knitr::include_graphics('../figures/tels-map.png')
```

@@

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
```

@@

## Datasets

@@

\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014. We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needleleaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively). We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package. Finally, we obtaied two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`. All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.<!-- The degrees of freedom and Frequentist confidence intervals were adjusted accordingly.--> The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).

@@

### Movement frequency and speed

@@

\small

@@

### Habitat selection

@@

\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.27), mean speed when moving (range: 0.42 -- 2.67 km/day), and mean distance travelled (range: 0.04 -- 0.62 km/day) even after post-stratifying to a sampling interval of one hour (Tab. \@ref(tab:gam-intercepts)). The inclusion of temperature in the model did not affect any of the three measures substantially when post-stratifying to $0^\circ$C. Grizzly bears tended to move least often ($P(M = 1) \approx 0.05$), while wolves and pumas moved most often ($P(M = 1) \ge 0.23$), and wolves had the highest mean speed when moving. Consequently, wolves traveled 3.0 to 15.6 times further per day than other mammals, on average.

* CHANGED FOREST RASTER TO USE ALL TREE TYPES (2025-01-17)
* NEED TO REMAKE HRSF FIGURES

***THIS IS FALSE***We did not find any consistent selections for or against forest cover, but all species exhibited stronger selection preferences for elevation and distance from water than for forest cover, and there was no consistent preference for forest cover across temperatures. In contrast, all species showed a strong preference for specific altitudes: wolves, pumas, and boreal caribou generally preferred elevations < 1 km above sea level, while elk, mountain goats, southern mountain caribou, and grizzly bears generally preferred elevations between 1 and 2 km above sea level. All species except for elk and grizzly bears preferred being as close to water as possible,

Overall, we did not detect any strong selections for or against forest cover, and this lack of preference was generally true irrespective of temperature. In contrast, we selection strength for forest cover did not vary strongly with temperature,

preferred resources

**CHECK TABLE 2: UPDATE AFTER REMOVING SCEK014 AND SCEK014b (2024-11-19 9:50)**

***UPDATE TABLES BASED ON NEW HRSFS***

@@

intercepts <- tibble(Species = SPECIES,
                     'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
                     'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
                     'E(S|M=1)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
                     'E(S|M=1,T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
                     'E(D)' = `P(M)` * `E(S|M=1)`,
                     'E(D|T)' = `P(M|T)` * `E(S|M=1,T)`) %>%
  mutate(Species = paste0('\\emph{', Species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .))

if(FALSE) {
  hist(intercepts$`P(M)`)
  hist(intercepts$`E(S|M=1)`)
  round(range(max(intercepts$`E(D)`) /
                intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]),
        1)

  intercepts %>%
    transmute(Species,
              p = round(`P(M|T)` - `P(M)`, 2),
              s = round(`E(S|M=1,T)` - `E(S|M=1)`, 2),
              d = round(`E(D|T)` - `E(D)`, 2))
}

intercepts %>%
  arrange(Species) %>%
  mutate(Species = gsub('angifer', '\\.', Species)) %>%
  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
  rename_with(\(.x) gsub('E', '\\\\mathbb E', .x)) %>%
  rename_with(\(.x) gsub('M=1', 'M\\\\!=\\\\!1', .x)) %>%
  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S | M = 1)$), and distance travelled ($\\mathbb E(D)$) as estimated by models without and with temperature, after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
```

@@

For most models, including smooth terms of temperature did not always improve within-sample prediction substantially (change in deviance explained: +0.2% -- + 17.9%), and there was relatively strong agreement between models with and without temperature (Fig. ***SX pred vs pred***), but including temperature resulted in substantially better fits for all models (all $\Delta$AIC $\le$ -326; Table \@ref(tab:temp-comparison)).

```{r temp-comparison}
setwd('..')
library('dplyr')
source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES`

temp_comp <- tibble(
  Model = c('P(moving)', 'Speed', rep('HRSF', 7)),
  Species = c('All', 'All', sort(as.character(SPECIES))),
  DDE = 
    c(10.7 - 10.2, 79.3 - 79.1, 10.0, 4.47, 17.6, 7.83, 5.89, 11.8, 17.9) %>%
    round(1) %>%
    paste0('%'),
  DAIC = c(-5927, -1682, -7200, -4305, -769, -945, -6284, -326, -920) %>%
    round())

@@

knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed), as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
```

@@

```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each speices' observed locations."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

### Predicted changes in movement during the current century

>>>>

---
title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs}
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection. We then leverage the estimated effects to predict how and where mammals will move in British Columbia (BC), Canada, throughout the 21^st^ century. We show that changes in temperature altered movement phenology, and that climate change will impact mammals' movement rates and habitat selection throughout the century, but that there are no common trends across all study species. Consequently, we suggest that species' responses be interpreted carefully and considered jointly when extrapolating the community-level consequences of climate change."
```

@@

\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behavior, habitat selection

@@

* A review of climate change effects on marine mammals in United States waters: Past predictions, observed impacts, current research and conservation imperatives https://doi.org/10.1016/j.ecochg.2022.100054
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* boars' sleep quantity, efficiency and quality are reduced on warmer days (https://doi.org/10.1098/rspb.2023.2115)

@@

<!-- movement is a convenient way to respond to change, but movement depends on temperature -->

\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Additionally, temperature also affects movement indirectly by through other drivers of movement behavior, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023]. But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024]. However, generalists' grater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to adapt within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021]. Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].

@@

Uncertainty in future conditions and their consequences on ecosystems presents a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behavior and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024]. In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare for and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [e.g., @tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [e.g., @zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future. Failing to understand how changes in temperature will affect mammalian movement behavior and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data thoughout a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behavior of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.

```{r bc-map, fig.cap="GPS telemetry data for the six species in this study.", out.width='0.75\\linewidth'}
knitr::include_graphics('../figures/tels-map.png')
```

@@

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
```

@@

## Datasets used

@@

\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014. We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively). We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package. Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`. All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).

@@

### Effects of temperature on movement rates

@@

\scriptsize

@@

### Effects of temperature on habitat selection

@@

\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)). Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$). When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s). Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.

@@

intercepts <- tibble(Species = SPECIES,
                     'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
                     'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
                     'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
                     'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
                     'E(D)' = `P(M)` * `E(S)`,
                     'E(D|T)' = `P(M|T)` * `E(S|T)`) %>%
  mutate(Species = paste0('\\emph{', Species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .))

if(FALSE) {
  hist(intercepts$`P(M)`)
  hist(intercepts$`E(S)`)
  round(range(max(intercepts$`E(D)`) / intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]), 1)

  intercepts %>%
    transmute(Species,
              p = round(`P(M|T)` - `P(M)`, 2),
              s = round(`E(S|T)` - `E(S)`, 2),
              d = round(`E(D|T)` - `E(D)`, 2))

  intercepts %>%
    arrange(desc(`E(D)`)) %>%
    mutate(displ_wolf = `E(D)`[grepl('Canis lupus', Species)] / `E(D)`,
           displ_wolf_t = `E(D|T)`[grepl('Canis lupus', Species)] / `E(D|T)`)
}

intercepts %>%
  arrange(Species) %>%
  mutate(Species = gsub('angifer', '\\.', Species)) %>%
  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
  rename_with(\(.x) gsub('E', '\\\\mathbb E', .x)) %>%
  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S)$), and distance travelled ($\\mathbb E(D)$) as estimated by models without and with temperature ($(... | T)$), after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
```

@@

At average temperatures, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig. ***SX pred vs pred***), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).

```{r temp-comparison}
setwd('..')
library('dplyr')
source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES` object

temp_comp <- tibble(
  Model = c('P(moving)', 'Speed', rep('HRSF', 7)),
  Species = c('All', 'All', sort(as.character(SPECIES))),
  DDE = 
    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
    round(1) %>%
    paste0('%'),
  DAIC = c(-5927, -1682, -6566, -4448, -813, -966, -5637, -342, -914) %>%
    round())

if(FALSE) {
  range(as.numeric(gsub('%', '', temp_comp$DDE)))
  range(as.numeric(gsub('%', '', temp_comp$DDE))[- c(1:2)])
  max(temp_comp$DAIC)
}

@@

knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
```

@@

Species' relative selection strength (RSS) for resources depended on temperature, and RSS was strongest for elevation and weakest for forest cover. Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou became more selective. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, cougars, and boreal caribou increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Estimated RSS values were generally most uncertain at extreme temperatures (Fig. ***BX***).

```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each speices' observed locations. RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

## Predicted changes in movement during the 21^st^ century
